---
title: "IASDT modelling workflow — 3. biotic data"
output: html_document
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  eval: false
format:
  html:
    self_contained: true
    mode: selfcontained
    embed-resources: true
    code-overflow: wrap
    code-link: true
    page-layout: full
    # toc-expand: true
    # toc-depth: 1
    # number-sections: true
    link-external-icon: true
    link-external-newwindow: true
vignette: >
  %\VignetteIndexEntry{IASDT modelling workflow — 3. biotic data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r}
#| include = FALSE,
#| eval = TRUE
library(kableExtra)
library(knitr)
library(tibble)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = FALSE, warning = FALSE,
  message = FALSE, dev = "ragg_png", dpi = 300, tidy = "styler",
  out.width = "100%", fig.show = "hold", echo = FALSE)
```

```{r}
#| eval = TRUE,
#| results = "asis"
c("<style>", readLines("style.css"), "</style>") %>% 
  cat(sep = "\n")
```

<br/><br/>

This article details the processing of biotic data within the `IASDT` modelling workflow. The biotic data are sourced from `GBIF`, `EASIN`, and `eLTER`; see below. All data preparation adheres to the <a href="https://doi.org/10.1038/sdata.2016.18" target="_blank" class="ll">FAIR</a> principles (Findable, Accessible, Interoperable, and Reusable) to ensure scientific integrity and reproducibility.

<br/><br/>

## IAS species list

The most recent checklist of naturalised terrestrial alien plant species of non-European origin was retrieved from <a href="https://floraveg.eu/" target="_blank" class="ll">FloraVeg.EU</a>. This checklist was standardized against the GBIF taxonomic backbone to ensure taxonomic consistency, resulting in a total of 1,319 species (September, 2025). The table below summarises the species count by taxonomic class and order. Occurrence data for these species were compiled from available sources detailed in subsequent sections.

```{r}
#| echo = FALSE,
#| message = FALSE,
#| warning = FALSE,
#| eval = TRUE
DT <- tibble::tribble(
  ~"class",         ~"order",            ~`# species`,
  "Liliopsida",     "Acorales",           2,
  "Liliopsida",     "Alismatales",        8,
  "Liliopsida",     "Arecales",           5,
  "Liliopsida",     "Asparagales",       74,
  "Liliopsida",     "Commelinales",       4,
  "Liliopsida",     "Liliales",           6,
  "Liliopsida",     "Poales",           163,
  "Liliopsida",     "Zingiberales",       7,
  "Lycopodiopsida", "Selaginellales",     1,
  "Magnoliopsida",  "Apiales",           15,
  "Magnoliopsida",  "Asterales",        148,
  "Magnoliopsida",  "Boraginales",       10,
  "Magnoliopsida",  "Brassicales",       13,
  "Magnoliopsida",  "Caryophyllales",   141,
  "Magnoliopsida",  "Celastrales",        4,
  "Magnoliopsida",  "Cornales",          10,
  "Magnoliopsida",  "Cucurbitales",       6,
  "Magnoliopsida",  "Dipsacales",        19,
  "Magnoliopsida",  "Ericales",          40,
  "Magnoliopsida",  "Escalloniales",      1,
  "Magnoliopsida",  "Fabales",           52,
  "Magnoliopsida",  "Fagales",           20,
  "Magnoliopsida",  "Garryales",          1,
  "Magnoliopsida",  "Gentianales",        9,
  "Magnoliopsida",  "Geraniales",         9,
  "Magnoliopsida",  "Gunnerales",         1,
  "Magnoliopsida",  "Lamiales",          63,
  "Magnoliopsida",  "Laurales",           1,
  "Magnoliopsida",  "Magnoliales",        1,
  "Magnoliopsida",  "Malpighiales",      41,
  "Magnoliopsida",  "Malvales",          10,
  "Magnoliopsida",  "Myrtales",          74,
  "Magnoliopsida",  "Oxalidales",        19,
  "Magnoliopsida",  "Piperales",          3,
  "Magnoliopsida",  "Proteales",          4,
  "Magnoliopsida",  "Ranunculales",      27,
  "Magnoliopsida",  "Rosales",          125,
  "Magnoliopsida",  "Sapindales",        23,
  "Magnoliopsida",  "Saxifragales",      38,
  "Magnoliopsida",  "Solanales",         51,
  "Magnoliopsida",  "Vitales",            9,
  "Pinopsida",      "Pinales",           46,
  "Polypodiopsida", "Cyatheales",         1,
  "Polypodiopsida", "Equisetales",        1,
  "Polypodiopsida", "Polypodiales",      12,
  "Polypodiopsida", "Schizaeales",        1)

knitr::kable(
  cbind(DT[1:23,], DT[24:46,]), format = "html", 
  align = "l", escape = FALSE) %>% 
  kableExtra::column_spec(c(1, 2, 4, 5), width_min = "1.4in")  %>% 
  kableExtra::column_spec(c(3, 6), width_min = "1.2in")  %>%
  kableExtra::kable_styling(
    c("basic", "condensed", "hover"), font_size = 16,
    position = "center", full_width = TRUE)
```

<br/><br/>

## GBIF

The `gbif_process()` function manages the retrieval, processing, and visualisation of the most current occurrence data for invasive alien species (IAS) from the Global Biodiversity Information Facility (<a href="https://www.gbif.org/" target="_blank" class="ll">GBIF</a>), totalling approximately 10.4 million occurrences as of September 2025. Occurrences deemed doubtful or exhibiting high spatial uncertainty are excluded to ensure data quality. Additionally, the function calculates the number of observations and occupied grid cells per species. Its primary output consists of gridded distribution maps for each species, delineating presence-only data across the reference grid.

<br/><br/>

## EASIN

The European Alien Species Information Network (<a href="https://easin.jrc.ec.europa.eu/" target="_blank" class="ll">EASIN</a>) serves as a web-based platform providing spatial data on approximately 14,000 alien species. The `easin_process()` function oversees the processing and visualisation of the latest IAS occurrence data sourced from EASIN. A checklist of vascular plants from EASIN’s database was retrieved via its API and standardized against the GBIF taxonomic backbone. Occurrence data for taxa corresponding to the IAS checklist were subsequently downloaded through the API. Of the 34 partner organizations contributing to EASIN (including GBIF), only non-GBIF data are incorporated into the models to avoid redundancy. Following processing and spatial filtering, the dataset includes over 427K observations for 465 IAS as of September 2025. Consistent with the GBIF output, the function generates gridded distribution maps for each species.

<br/><br/>

## eLTER

The Integrated European Long-Term Ecosystem, Critical Zone, and Socio-Ecological Research Infrastructure (<a href="https://elter-ri.eu/" target="_blank" class="ll">eLTER</a>) comprises a network of sites dedicated to collecting ecological data for long-term research across the European Union. Cleaned standardised data encompassed 4K observations for 109 IAS (September 2025). The `elter_process()` function processed standardised eLTER data in a format ready to be used in the models.

<br/><br/>

## Species distribution

The `naps_process()` function integrates and visualises species distribution data sourced from `GBIF`, `EASIN`, and `eLTER`. It combines pre-processed datasets and generates presence-absence rasters for use in the species distribution models.

<hr class="hr1">

<span style="font-size: 1.2em; line-height: 0.8;">
<b>Previous articles:</b><br/>
<tab>&#8608;<tab><a href="workflow_1_overview.html" class="ll">1. Overview</a><br/>
<tab>&#8608;<tab><a href="workflow_2_abiotic_data.html" class="ll">2. Processing abiotic data</a><br/>
<b>Next articles:</b><br/>
<tab>&#8608;<tab><a href="workflow_4_model_fitting.html" class="ll">4. Model fitting</a><br/>
<tab>&#8608;<tab><a href="workflow_5_model_postprocess.html" class="ll">5. Model post-processing</a><br/>
</span>
